---
title: "Better Household Living Conditions Improve Life Satisfaction"
author: "Rachel Oh, Yang Shang"
date: "19/10/2020"
abstract: "This report describes the effects of household living conditions on life satisfaction. It is found that better household living conditions do improve life satisfaction to a certain degree. These results are important as it is fundamental that people know what makes them satisfied with their lives because no human would wish to have an unsatisfying life."
output: pdf_document
---


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#### Packages installation ####
#install.packages("tidyverse")
#### Workspace set-up ####
library(tidyverse)
```

# Introduction

In our studies, we established an interest about the relationship between people's life satisfaction and their living conditions. Resourced from Canadian GSS in 2017, our team is going to describe an individual's living conditions from the following aspects:

1. Household type: what type of residence is the respondent living in (house, apartment, etc.)?
2. Household size: how many residents (relatives and non-relatives), including the respondent, are living in the residence?
3. Ownership of the residence: is the respondent owning or renting the residence?
4. Cohabitation: Is the respondent living with a partner (married or not married)
5. Children in household: Is there any child (relatives and non-relatives) living in the residence?
The detail of the above factors will be further described and explained in Section "Model".

Our prediction to the model is that: 1. people live in houses have higher life satisfaction; 2. the larger the household size, the higher the life satisfaction; 3. people owning the residence have higher life satisfaction; 4. people live with their partners have higher life satisfaction; 5. people live with children have higher life satisfaction.

Full code and data supporting this analysis is available at: https://github.com/hyunbani/household-living-conditions-and-life-satisfaction

# Data

## General Description of GSS

Statistics Canada's General Social Survey (GSS) is an annual survey with statistical information on the living conditions and well-being of people living in Canada. In this report, we are using the data collected in 2017. The main focus of that year's survey is "Family".

The Survey is divided into 86 minor sections, with around 200 questions in total, to describe the respondent's general identity, marriage, family relationships and identity, child care, social involvements, living conditions, career, and religion.

People of interest were those aged at 15 and older, living in all 10 provinces in Canada. However, it would be costly and impractical to survey each and every household in Canada, so random sampling was chosen as a method to determine the sample of 43,000 people, which reflects the population. In order to ensure that the sample is an accurate reflection of the population as a whole, the survey results from all sampled households were collected, and only one eligible person per household was interviewed.

## Collection of GSS Data

The 2017 GSS was performed through telephone interviews. People who did not answer the telephone calls were excluded from the sample. However, there still exist some non responses for some of the questions being asked, and these are recorded as 'NA' in the dataset. 

Since the answers are kept strictly confidential, the respondents must have provided the answers in a more honest manner, leading to an increase in accuracy of data information. However, as there are about 200 questions in total, even if some can be validly skipped according to respondent's status, it still took a lot of time to complete one servey. Besides, many people tend to refuse the phone calls. Thus, we believed that telephone interview might not be the most efficient collecting method. 

## About Our Studies

The team's interest of study is the relationship between life satisfaction and living conditions. We extract colume "feelings_life" to represent life satisfaction. We also extract columes "own_rent", "hh_type", "hh_size", "lives_with_partner", and "children_in_household" to describe household living conditions.

Our response variable "feelings_life" is a discrete numerical variable measured on a scale of intergers from 0 to 10, with 10 being the highest satisfaction level and 0 being the lowest. The team was previously interested in using data "self_rated_mental_health", which use categorical variables "poor", "fair", "good", "very good", "exellent" to describe the respondent's mental conditions. However, we experience difficulties creating and summerizing the model using this variable, even if we give it numerical expression. Then we switch to the current variable "feelings_life". We believe that it is an upper level data included the effect of respondent's mental health.

Figure 1 below is a bar graph showing the proportion of respondents' life satisfaction.


```{r plot, echo=FALSE, message=FALSE, results='hide'}
gss_household <- read_csv("gss_cleaned.csv")
gss_household <- na.omit(gss_household)
gss_household <- gss_household %>%
  mutate(own_rent = case_when(
    own_rent == "Owned by you or a member of this household, even if it i..." ~ "Owned",
    own_rent == "Rented, even if no cash rent is paid" ~ "Rented"
  ))

gss_household %>%
  ggplot(mapping = aes(x=feelings_life)) + 
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  ggtitle("Figure1: Proportion of Respondents' Life Satisfaction")+
  xlab("Life Satisfaction Level")+
  ylab("Proportion")+
  scale_x_continuous(breaks = seq(0,10,1))

summary(gss_household$feelings_life)
```

The graph in figure 1 shows that about 40% of the respondents rated their life satisfaction to 9 and 10, about 45% of the respondents rated 7 and 8 (with 30% in total rated 8), 10% rated 5 and 6, and less than 5% rated 4 and below. The median is 8 and mean life satisfaction is 8.101. We are pretty confident to say that most of the respondents have a pretty good attitudes toward life.

Variable "own_rent" illustrates if the respondents owned or rented their current residence. We believe that it is related to respondents' income level. However, factors like family history also affects respondents' ownership to a residence.

Variable "own_rent" illustrates if the respondents owned or rented their current residence. The following figure shows the proportion of residence ownership of the respondents.

Variable "hh_type" shows the type of residence the respondents were living in, including "high rise apartment", "low rise apartment", "single detached house", and "other". The team has simplified the variable to two parameters, "single detached house" and "other" representing the other 3 residence types.

Variable "hh_size" is a categorical variable specified by integers from 1 to 6 showing the size of the household. "6" is specially representing a household of 6 or more people. In our model, we will treat this as a discrete numeric variable.

Variable "lives_with_partner" is a categorical variable showing if the respondent is living with a partner (married or not married). The variable includes some effect of the respondents' status of marriage. However, the variable doesn't measure the effect of parents in a household as some respondents probably lives together with their parents.

Variable "children_in_household" is a categorical variable showing how many children are living in the household. Non-relative children are also counted. The original data is classified by "no child", "one child", "two children", and "three or more children". In our model, we simplified it to just measure the effects of having child or not inside the household.

For the test set the team is using in the model, we have created some dummy variables for the data set. The default setting of a respondent is: "rent the residence", "lives in other types of residence", "lives without partner", "lives without child in household".

We have also modified names to the variables. The explanation of each explanatory variable is as follows:\
1) 'Household Size' - the number of people each respondent lives with\
2) 'Owned' - 1 if a respondent owns the house and 0 if he/she doesn't\
3) 'Partner' - 1 if a respondent lives with a partner and 0 if he/she doesn't\
4) 'No Child' - 1 if no child lives in the household with the respondent and 0 if he/she does\
5) 'Single House' - 1 if respondent's house type is single detached house and 0 if it's not 

Table 1 below shows the first ten rows of the data with the variables of interest.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# this is needed to form a table
library(knitr)
library(ggplot2)

# gss_data.csv (reduced and cleaned raw data) is brought from the outputs folder and was created in Data.R
gss_data = read.csv("outputs/gss_data.csv")

# forming a table showing the first ten rows of the above data
knitr::kable(gss_data[0:10, ], col.names = c("Case", "Life Satisfaction", "Household Size", "Owned", "Partner", "No Child", "Single House"), caption = "First 10 Rows of Raw Data")
```


# Model

We are interested in explaining whether a person's life satisfaction improves when he/she has better household living conditions, based on household size, house ownership, whether he/she has a partner, no child, and lives in a single detached house. In order to do so, we are going to fit a multiple linear regression model.

Multiple linear regression is a statistical technique that uses several explanatory variables to predict the outcome of a response variable. The goal of it is to model the linear relationship between the explanatory (independent) variables and response (dependent) variables.

As discussed in the data section, the dependent variable is Life Satisfaction and the explanatory variables are Household Size, Owned, Partner, No Child and Single House.

Formula and calculation of this model is as follows:
$$y_i=\beta_0+\beta_1{x_i}_1+\beta_2{x_i}_2+\beta_3{x_i}_3+\beta_4{x_i}_4+\beta_5{x_i}_5+\epsilon$$
where, for $i=20331$ observations:\
$y_i=$ dependent variable\
$x_i=$ explanatory variables\
$\beta_0=$ y-intercept (constant term)\
$\beta_1...\beta_5=$ slope coefficients for each explanatory variable\
$\epsilon=$ the model's error term (also known as the residuals)

```{r, echo=FALSE, warning=FALSE}
gss_model<-lm(life_satisfaction~household_size+house_owned+live_with_partner+no_child+house,data=gss_data)
```


# Results

```{r, echo=FALSE, warning=FALSE}
gss_summary <- summary(gss_model)$coefficients
row.names(gss_summary)<-c("(Intercept)","Household Size","Owned","Partner","No Child","Single House")
knitr::kable(gss_summary,digits=2, caption = "Summary of the Model")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
relweights <-
function(fit,...){                         
  R <- cor(fit$model)   
  nvar <- ncol(R)          
  rxx <- R[2:nvar, 2:nvar] 
  rxy <- R[2:nvar, 1]      
  svd <- eigen(rxx)        
  evec <- svd$vectors                           
  ev <- svd$values         
  delta <- diag(sqrt(ev))  
  lambda <- evec %*% delta %*% t(evec)        
  lambdasq <- lambda ^ 2   
  beta <- solve(lambda) %*% rxy           
  rsquare <- colSums(beta ^ 2)                   
  rawwgt <- lambdasq %*% beta ^ 2    
  import <- (rawwgt / rsquare) * 100 
  import <- as.data.frame(import)
  row.names(import)<-names(fit$model[2:nvar])
  names(import)<-"Weights"
  import<-import[order(import),1,drop=FALSE]
  dotchart(import$Weights,labels=c("No Child","Partner","Household Size","Single House","Owned"),
           xlab="% of R-Square",pch=19,
           main="Figure 2: Relative Importance of Predictor Variables",
           sub=paste("Total R-Square=",round(rsquare,digits=3)),
           ...)
  return(import)
}
relweights(gss_model)
```


# Discussion

From Table 2, We see that the p-values (represented as Pr(>|t|)) for all explanatory variables are less than 0.10. From this, we know that all these factors increase the life satisfaction level with a significance level of 0.10. It means that you will be more satisfied with your life if you have more people living in your household, own the house, live with a partner, live with no child, and live in a single detached house.

However, the significance for each explanatory variable is different on the response variable. If you look at Figure 2, you will see that whether you own the house or not has the biggest impact on your life satisfaction level. The variable with the least impact is having no child and this can also be seen through the graph. 

The error of our model seems to have the greatest effect on our model. The effect is greater than the sum of all the influences of our variables in the model we have. We are still confident that household living conditions has a noticeable impact on respondents' life satisfaction. However, we suspect that there is other factor other than living conditions that significantly affect respondents' life satisfaction. Focuses like religious effects, education effects, career effects and/or a more detailed version of living conditions can be considers.

# References

Canadian General Social Surveys (2017). General social survey on Family (cycle 31). URL https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&Id=335816

Hadley Wickham and Evan Miller (2020). haven: Import and Export 'SPSS', 'Stata' and 'SAS' Files.
  R package version 2.3.1. URL https://CRAN.R-project.org/package=haven

  Hadley Wickham. ggplot2: Elegant Graphics for Data Analysis.
  Springer-Verlag New York, 2016.
  
Jacob Kaplan (2020). fastDummies: Fast Creation of Dummy (Binary) Columns and Rows from
  Categorical Variables. R package version 1.6.2. URL https://CRAN.R-project.org/package=fastDummies

JJ Allaire and Yihui Xie and Jonathan McPherson and Javier Luraschi and Kevin Ushey and Aron
  Atkins and Hadley Wickham and Joe Cheng and Winston Chang and Richard Iannone (2020). rmarkdown:
  Dynamic Documents for R. R package version 2.3. URL https://rmarkdown.rstudio.com.

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686. URL
  https://doi.org/10.21105/joss.01686

Yihui Xie and J.J. Allaire and Garrett Grolemund (2018). R Markdown: The Definitive Guide.
  Chapman and Hall/CRC. ISBN 9781138359338. URL https://bookdown.org/yihui/rmarkdown.
  
  Yihui Xie (2020). knitr: A General-Purpose Package for Dynamic Report Generation in R. R package
  version 1.30.

  Yihui Xie (2015) Dynamic Documents with R and knitr. 2nd edition. Chapman and Hall/CRC. ISBN
  978-1498716963

  Yihui Xie (2014) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden,
  Friedrich Leisch and Roger D. Peng, editors, Implementing Reproducible Computational Research.
  Chapman and Hall/CRC. ISBN 978-1466561595
  